---
title: "Univariate mixture clustering with the DPM engine"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Univariate mixture clustering with the DPM engine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette demonstrates Bayesian univariate Gaussian mixture clustering
with **nimix**'s Dirichlet Process Mixture (DPM) engine, built on NIMBLE's
Chinese Restaurant Process (`dCRP`), using **official statistics**: the
`wdi2022` dataset shipped with the package (World Bank World Development
Indicators, 2022, CC BY 4.0; see `?wdi2022`). The same workflow applies
directly to other official-statistics sources, e.g. BPS regional indicators.

> MCMC chunks use `eval = FALSE` so the vignette builds quickly under CRAN
> time limits; the printed results below are from an actual run. Run the
> chunks interactively to reproduce them.

## The data: national income across 207 countries

GDP per capita is famously multimodal on the log scale -- countries group into
latent "development regimes" rather than forming one homogeneous population.

```{r data, eval = TRUE}
library(nimix)
data(wdi2022)
ly <- log(wdi2022$gdp_pc)
summary(ly)
hist(ly, breaks = 30, main = "log GDP per capita, 2022", xlab = "log US$")
```

## Fit: how many income regimes?

`method = "dpm"` estimates the number of occupied components. `K_max` is only
a truncation level (computational headroom), not a prior belief about K.

```{r fit}
fit <- nimixClust(ly, K_max = 10,
                  mcmcControl = list(niter = 4000, nburnin = 1500),
                  seed = 1)
fit <- relabel(fit)   # always relabel before summarising (label switching)
summary(fit)
```

On the 2022 data this finds a **two-regime** structure:

```
#> modalK = 2
#>   weight mu_mean
#> 1  0.554   7.966    # ~ US$ 2,900  (lower/middle-income regime)
#> 2  0.446   9.964    # ~ US$ 21,000 (high-income regime)
```

The posterior on the number of clusters concentrates on 2-3, with the usual
DPM tail of small transient components. If the fit warns that the truncation
level was reached in a non-negligible share of iterations, increase `K_max`
and re-run -- a larger truncation only adds headroom.

## Multi-chain convergence check

Multiple chains from dispersed starts cost almost nothing extra: the compiled
model is reused across chains, and `summary()` then reports cross-chain
split-Rhat / ESS on label-invariant quantities.

```{r chains}
fit3 <- nimixClust(ly, K_max = 10,
                   mcmcControl = list(niter = 4000, nburnin = 1500,
                                      nchains = 3),
                   seed = 1)
summary(relabel(fit3))
```

## Sanity check with a fixed K

`method = "fixedk"` fits a classic finite mixture with known K -- useful as a
baseline or for model comparison across a few K values.

```{r fixedk}
fit2 <- nimixClust(ly, K = 2, method = "fixedk",
                   mcmcControl = list(niter = 4000, nburnin = 1500),
                   seed = 1)
summary(relabel(fit2))
```
