---
title: "Regime switching in time series: the hidden-Markov engine"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regime switching in time series: the hidden-Markov engine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

An ordinary mixture assumes each observation picks its component
independently. Time series usually violate that: a market sits in a
low-volatility regime for months, then switches; a physiological signal
alternates between states. `method = "hmm"` replaces the independent
allocation prior with a first-order Markov chain,
$z_t \mid z_{t-1} \sim \mathrm{Cat}(P[z_{t-1},\,])$, so *persistence itself*
is estimated -- the diagonal of the transition matrix $P$.

Two implementation facts shape everything downstream:

1. **The state path is marginalised out of the likelihood** by the forward
   algorithm, so the MCMC samples only the continuous parameters. Measured on
   the gate prototype: 4000 iterations in about a second at $T = 300$, with
   min ESS/sec 456 against 144 for the equivalent model that samples the
   latent states directly.
2. **Allocation draws are recovered afterwards, exactly**, by forward-filter
   backward-sampling (FFBS) per retained draw. That is why everything you
   already know -- `relabel()`, `psm()`, `binderPartition()`, the plots, the
   bayesplot adaptors -- works on an HMM fit unchanged.

> MCMC chunks use `eval = FALSE` (CRAN time limits); printed results are from
> an actual run.

## Simulate two persistent regimes

```{r}
library(nimix)

set.seed(11)
P  <- rbind(c(0.95, 0.05),    # regime 1: very persistent
            c(0.10, 0.90))    # regime 2: persistent
mu <- c(-2, 2); sg <- c(0.6, 0.8)
z  <- integer(300); z[1] <- 1
for (t in 2:300) z[t] <- sample(1:2, 1, prob = P[z[t - 1], ])
y  <- rnorm(300, mu[z], sg[z])
```

The data order **is** the time order -- that is what the Markov prior binds
to.

## Fit

`K` is the number of states (fixed, like `method = "fixedk"`); current
emission families are univariate Gaussian, Student-t (heavy-tailed regimes
such as financial returns), Poisson (regime-switching counts), and the
neo-normal skewed families MSNBurr, MSNBurr-IIa and GMSNBurr, with further
families following the gated roadmap.

```{r}
fit <- nimixClust(y, K = 2, method = "hmm",
                  mcmcControl = list(niter = 4000, nburnin = 1500,
                                     nchains = 2),
                  seed = 1)

fit <- relabel(fit)
fit@relabeled$summary[, c("mu_mean", "s2_mean")]
#>   mu_mean s2_mean
#> 1   -1.99   0.432
#> 2    2.12   0.868      # truth: mu -2, 2; s2 0.36, 0.64
```

The estimated persistence is read straight off the transition draws:

```{r}
mean(fit@mcmcSamples[, "P[1, 1]"])
#> [1] 0.965                          # truth 0.95; 95% CI [0.936, 0.985]
mean(fit@mcmcSamples[, "P[2, 2]"])
#> [1] 0.896                          # truth 0.90
```

## Two decodings, one distinction worth knowing

`viterbiPath()` returns the single *jointly* most probable state sequence
under the Markov prior; the FFBS allocation draws give the *marginal*
posterior of each $z_t$, which `binderPartition()` and per-time MAP
summarise. On well-separated regimes they agree; they part exactly where the
state is genuinely ambiguous.

```{r}
zv <- viterbiPath(fit)
sum(diff(zv) != 0) + 1
#> [1] 15                             # 15 regime runs -- matches the truth

zmap <- apply(fit@clusterAllocation, 2,
              function(v) as.integer(names(which.max(table(v)))))
mean(zv == zmap)
#> [1] 1                              # full agreement here
```

Because the FFBS draws populate `clusterAllocation`, the label-free partition
tools need no adaptation:

```{r}
bp <- binderPartition(fit)
bp$K
#> [1] 2
S <- psm(fit)                        # 300 x 300 posterior similarity
```

Plotting `image(S)` is a useful regime diagnostic: block structure along the
diagonal is persistence made visible.

## Diagnostics and predictive checks

The invariant draws array works as for any engine (the concentration
parameter `alpha` does not exist here and is dropped automatically):

```{r}
da <- drawsArray(fit)                # 2500 x 2 chains x {K, entropy}
# bayesplot::mcmc_trace(da)
```

`ppCheck()` and `ppcData()` also work, with one honest nuance: replicated
series are drawn **conditionally on each draw's sampled state path**, not by
re-simulating the Markov chain. That is the right check for "do the
emissions fit, given the regimes"; it does not test the transition structure
itself, for which the run-length comparison above is the more direct probe.

```{r}
pd <- ppcData(fit, ndraws = 20)      # list(y, yrep), yrep 20 x 300
# bayesplot::ppc_dens_overlay(pd$y, pd$yrep)
```

## What to watch out for

- **Order matters.** The Markov prior binds to the data order; shuffled data
  silently means an i.i.d. mixture fitted expensively.
- **State labels can switch**, like any mixture. `relabel()` handles the
  component summaries; `viterbiPath()` and the partition tools are
  label-free by construction.
- **Over-specifying `K` is safe but diluting.** Extra states go (nearly)
  empty -- the engine's tests assert no corruption occurs -- but transitions
  into barely-used states are poorly informed; keep `K` at the number of
  regimes you can defend.

## References

Rabiner, L. R. (1989). A tutorial on hidden Markov models and selected
applications in speech recognition. *Proceedings of the IEEE* 77, 257--286.

Frühwirth-Schnatter, S. (2006). *Finite Mixture and Markov Switching
Models*. Springer.
