---
title: "Multivariate mixture clustering with the DPM engine"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multivariate mixture clustering with the DPM engine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette demonstrates **multivariate** Gaussian mixture clustering on the
DPM engine. From the user's point of view the only change from the univariate
case is that `data` is a numeric **matrix** (one row per observation, one
column per dimension). The example clusters countries in the `wdi2022`
official-statistics dataset (`?wdi2022`) on two development indicators
jointly.

> MCMC chunks use `eval = FALSE` (CRAN time limits); printed results are from
> an actual run.

## The data: income and longevity jointly

```{r data, eval = TRUE}
library(nimix)
data(wdi2022)
Y <- cbind(log_gdp  = log(wdi2022$gdp_pc),
           life_exp = wdi2022$life_exp)
plot(Y, main = "207 countries, 2022", xlab = "log GDP per capita",
     ylab = "life expectancy (years)")
```

## The model

Each component is a multivariate normal with a conjugate
Normal-Inverse-Wishart cluster base measure (`dmnorm(cov=)` + `dinvwish`),
data-scaled by default. NIMBLE's native conjugate CRP samplers are used --
nimix deliberately does not reimplement them.

## Fit

```{r fit}
fit <- nimixClust(Y, K_max = 8,
                  mcmcControl = list(niter = 4000, nburnin = 1500),
                  seed = 1)
fit <- relabel(fit)
summary(fit)
```

On the 2022 data the dominant structure is a clearly separated
**high-income / high-longevity** cluster plus overlapping developing-country
clusters:

```
#> modalK = 4
#>   component weight   mu_1   mu_2      # (log GDP pc, life expectancy)
#> 3         3  0.309 10.363 80.302     # rich, long-lived cluster
#> 4         4  0.444  8.087 69.402     # large developing cluster
#> 1         1  0.204  8.570 71.788     # overlapping middle cluster
#> 2         2  0.043  8.556 69.914     # small transient component
```

Overlapping components with small weights are expected in a DPM; report the
dominant components and, when a hard partition is needed, use the MAP
allocation:

```{r map}
alloc <- apply(fit@clusterAllocation, 2L, function(z)
  as.integer(names(which.max(table(z)))))
plot(Y, col = alloc, pch = 19,
     main = "MAP cluster allocation")
```

## Fixed-K comparison

```{r fixedk}
fit3 <- nimixClust(Y, K = 3, method = "fixedk",
                   mcmcControl = list(niter = 4000, nburnin = 1500),
                   seed = 1)
summary(relabel(fit3))
```
