---
title: "Mixtures of linear regressions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixtures of linear regressions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`nimixReg()` fits a Bayesian mixture of Gaussian linear regressions: each
component is \(y \sim N(x^\top \beta_k, \sigma^2_k)\) with a conjugate
Normal-Inverse-Gamma cluster prior. The number of regimes can be inferred
(`method = "dpm"`) or fixed (`method = "fixedk"`).

This vignette uses the **Preston curve** -- the classic official-statistics
relationship between national income and life expectancy -- on the `wdi2022`
dataset (`?wdi2022`).

> MCMC chunks use `eval = FALSE` (CRAN time limits); printed results are from
> an actual run.

## The data

```{r data, eval = TRUE}
library(nimix)
data(wdi2022)
df <- data.frame(life = wdi2022$life_exp,
                 lgdp = log(wdi2022$gdp_pc))
plot(df$lgdp, df$life, xlab = "log GDP per capita",
     ylab = "life expectancy", main = "Preston curve, 2022")
```

## Does the Preston curve have latent regimes?

```{r fit}
fit <- nimixReg(life ~ lgdp, df, K_max = 6, method = "dpm",
                mcmcControl = list(niter = 5000, nburnin = 2000),
                seed = 1)
fit <- relabel(fit)
summary(fit)
```

An instructive result: on log income the DPM concentrates on a **single
regime** --

```
#> modalK = 1
#>   (Intercept)  lgdp
#> 1      36.782 4.067      # life ~ 36.8 + 4.07 x log GDP pc
```

roughly four extra years of life expectancy per log-dollar of income. The
lesson is that a nonparametric mixture does **not** invent regimes: when one
regression line (with its residual spread) explains the data, the posterior
collapses to it. Estimated K always depends on the assumed component form --
heterogeneity here lives in the residuals, not in distinct slopes.

Even when two components are *offered* (fixed K = 2), one simply empties:

```{r fixedk}
fit2 <- nimixReg(life ~ lgdp, df, K = 2, method = "fixedk",
                 mcmcControl = list(niter = 5000, nburnin = 2000),
                 seed = 1)
summary(relabel(fit2))
#> one occupied component; weight ~ 1 on the Preston line
```

## When regimes do exist

With genuinely regime-structured data the same call recovers them. A
simulated benchmark (two latent groups with slopes +2 / -2, mixed 80/20):

```{r sim}
set.seed(107)
n <- 250; x <- runif(n, -3, 3)
grp <- rep(1:2, c(200, 50))
y <- ifelse(grp == 1, 2 * x, -2 * x) + rnorm(n, 0, 0.7)
fs <- nimixReg(y ~ x, data.frame(y = y, x = x), K_max = 8,
               mcmcControl = list(niter = 5000, nburnin = 2000), seed = 1)
summary(relabel(fs))    # recovers slopes near +2 and -2, weights near 0.8/0.2
```

## Reading the error variance

One number deserves a caveat: the component error variance `s2`. Its prior is
centred on the residual variance of a *global* OLS fit, which ignores the
mixture — so for separated components it measures the spread *between*
components as much as the spread within one, and `s2` comes out too large.

The bias is largest exactly where mixtures earn their keep. At 150
observations per component it is about 1.0x for overlapping components and
4.7x for well-separated ones; against sample size it runs ~9x at 25
observations per component and fades to 1.0x by 5000. It is in the safe
direction — predictive intervals come out wider, not narrower — and the
coefficients are untouched.

The conservatism is deliberate: it also regularises against splitting when
`K` is over-specified. Shrinking it makes an over-specified fit occupy more
components than it should. So it stays the default — but if you know the
within-component scale, say so:

```{r priorscale, eval = FALSE}
nimixReg(y ~ x, df, K = 2, method = "fixedk",
         prior = list(s2Guess = 0.3))   # your prior mean for s2
```

On a simulated benchmark with a true `s2` of 0.25, that moved the estimate
from 0.78 to 0.36 while leaving the slopes where they were. The override
asks for an absolute value on purpose: a multiplier on the automatic scale
could only ever mean "a fraction of a quantity that measures the wrong
thing" — a knob rather than a statement.

Tightening the prior is safe with the DPM, and with `fixedk` when `K` is
right. It is *not* safe when `K` is over-specified: on the same data with
`K = 4` against two true components, the default kept two components
occupied while a tenfold tighter prior filled a third. If you are unsure
about `K`, leave the default alone or use the DPM.

### Recipe: two-stage empirical Bayes

If you need the scale right and have no external knowledge, let the first fit
tell you. Its *allocation* is reliable even where its `s2` is not — which is
exactly what makes this work:

```{r ebrecipe, eval = FALSE}
fit1 <- relabel(nimixReg(y ~ x, df, K = 2, method = "fixedk"))

z    <- binderPartition(fit1)$partition          # MAP component per point
cf   <- fit1@relabeled$summary
Xm   <- model.matrix(y ~ x, df)
res  <- df$y - rowSums(Xm * as.matrix(cf[z, c("(Intercept)", "x")]))
s2hat <- sum(res^2) / (nrow(df) - 2 * ncol(Xm)) # within-component residual var

fit2 <- nimixReg(y ~ x, df, K = 2, method = "fixedk",
                 prior = list(s2Guess = s2hat))
```

It costs a second compile-and-run, which is why it is not the default. Note
the same caveat as above: only reach for it when `K` is not over-specified.

## Prediction

```{r predict}
newd <- data.frame(lgdp = log(c(1000, 10000, 50000)))
predict(fit, newdata = newd)
```

Heavy-tailed residuals (`distribution = "studentt"` / `"normalgamma"`) and
multivariate responses (`cbind(y1, y2) ~ x`) use the same interface.

## Grouped data: a random intercept

When observations are nested in groups (regions, individuals, batches), a
shared group offset $b_{g(i)} \sim N(0, \tau^2)$ can be added to every
component's linear predictor with `random = ~ group`:

```{r re, eval = FALSE}
# grouped data with 12 groups; component slopes +2 / -2
reg <- nimixReg(y ~ x, df, random = ~ g, K = 2, method = "fixedk",
                mcmcControl = list(niter = 3000, nburnin = 1200), seed = 1)
sort(relabel(reg)@relabeled$summary$x)
#> [1] -1.93  1.97                    # slopes recovered with the offset present
mean(reg@mcmcSamples[, "tauRE"])       # the between-group SD, tau
```

Two things are worth knowing. The offsets use a **sum-to-zero**
parameterisation, because with free $b$ the component intercepts and the mean
of $b$ form a translation ridge that mixes badly (a prototype measured
`cor` = -0.979 and a minimum ESS of 25 in 2500 draws; the constraint restored
it to ~205). So the reported `b` are centred, and the component intercepts
absorb the group mean. And the exact conjugate coefficient sampler is
extended to condition on the current offsets, which keeps the fit
scale-equivariant in the predictors even with the random effect active.

Reading `tauRE` needs one caution. It estimates the spread of *your* group
offsets, and with few groups that spread is itself a lottery: twelve offsets
drawn from a population with $\sigma = 0.8$ have a realized standard
deviation anywhere between about 0.5 and 1.1. So an apparent "miss" against
a population value is usually the draw, not the estimator -- judge `tauRE`
against the groups you actually have.

Groups need not be balanced across components. With groups sitting almost
entirely in one component (a case that could plausibly let the offsets
absorb a component intercept), the coefficients still separate cleanly and
the offsets stay identified; the test suite asserts this.

### Random slopes

`random = ~ x | g` lets the slope of `x` vary by group as well (the intercept
comes along, as in `lme4`'s `(x|g)`):

```{r reslope, eval = FALSE}
reg <- nimixReg(y ~ x, df, random = ~ x | g, K = 2, method = "fixedk",
                mcmcControl = list(niter = 3000, nburnin = 1200), seed = 1)
```

There is an obvious worry here: components in a mixture of regressions are
distinguished *by* their slopes, so group-varying slopes might fight with
component identification. A prototype pushed this deliberately hard --
group slope deviations spanning $-2.05$ to $1.43$ against a component
separation of 4, so one group's effective slope landed exactly between the
two components -- and it held: group slopes recovered at `cor` = 0.997 and
component allocation at 95.3% accuracy.

The same sum-to-zero logic applies, so the component slopes absorb
`mean(s_g)` and the reported `sRE` are centred. The two sets of offsets get
independent priors; that was checked against correlated truth
($\rho = 0.92$), where the independent model still recovered both and the
correlation reappeared in the estimates anyway. With very few observations
per group the prior does more work and a correlated prior would matter more.

Both work with heavy-tailed residuals too — `distribution = "studentt"`
accepts the same `random =` argument, which is often what grouped data with
occasional outliers actually needs.

Current scope: `method = "fixedk"`, `distribution = "normal"` or
`"studentt"`, one grouping factor, random intercept and one random slope.
Further families follow the gated roadmap.
