---
title: "Mixtures of linear regressions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixtures of linear regressions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`nimixReg()` fits a Bayesian mixture of Gaussian linear regressions: each
component is \(y \sim N(x^\top \beta_k, \sigma^2_k)\) with a conjugate
Normal-Inverse-Gamma cluster prior. The number of regimes can be inferred
(`method = "dpm"`) or fixed (`method = "fixedk"`).

This vignette uses the **Preston curve** -- the classic official-statistics
relationship between national income and life expectancy -- on the `wdi2022`
dataset (`?wdi2022`).

> MCMC chunks use `eval = FALSE` (CRAN time limits); printed results are from
> an actual run.

## The data

```{r data, eval = TRUE}
library(nimix)
data(wdi2022)
df <- data.frame(life = wdi2022$life_exp,
                 lgdp = log(wdi2022$gdp_pc))
plot(df$lgdp, df$life, xlab = "log GDP per capita",
     ylab = "life expectancy", main = "Preston curve, 2022")
```

## Does the Preston curve have latent regimes?

```{r fit}
fit <- nimixReg(life ~ lgdp, df, K_max = 6, method = "dpm",
                mcmcControl = list(niter = 5000, nburnin = 2000),
                seed = 1)
fit <- relabel(fit)
summary(fit)
```

An instructive result: on log income the DPM concentrates on a **single
regime** --

```
#> modalK = 1
#>   (Intercept)  lgdp
#> 1      36.782 4.067      # life ~ 36.8 + 4.07 x log GDP pc
```

roughly four extra years of life expectancy per log-dollar of income. The
lesson is that a nonparametric mixture does **not** invent regimes: when one
regression line (with its residual spread) explains the data, the posterior
collapses to it. Estimated K always depends on the assumed component form --
heterogeneity here lives in the residuals, not in distinct slopes.

Even when two components are *offered* (fixed K = 2), one simply empties:

```{r fixedk}
fit2 <- nimixReg(life ~ lgdp, df, K = 2, method = "fixedk",
                 mcmcControl = list(niter = 5000, nburnin = 2000),
                 seed = 1)
summary(relabel(fit2))
#> one occupied component; weight ~ 1 on the Preston line
```

## When regimes do exist

With genuinely regime-structured data the same call recovers them. A
simulated benchmark (two latent groups with slopes +2 / -2, mixed 80/20):

```{r sim}
set.seed(107)
n <- 250; x <- runif(n, -3, 3)
grp <- rep(1:2, c(200, 50))
y <- ifelse(grp == 1, 2 * x, -2 * x) + rnorm(n, 0, 0.7)
fs <- nimixReg(y ~ x, data.frame(y = y, x = x), K_max = 8,
               mcmcControl = list(niter = 5000, nburnin = 2000), seed = 1)
summary(relabel(fs))    # recovers slopes near +2 and -2, weights near 0.8/0.2
```

## Prediction

```{r predict}
newd <- data.frame(lgdp = log(c(1000, 10000, 50000)))
predict(fit, newdata = newd)
```

Heavy-tailed residuals (`distribution = "studentt"` / `"normalgamma"`) and
multivariate responses (`cbind(y1, y2) ~ x`) use the same interface.

## Grouped data: a random intercept

When observations are nested in groups (regions, individuals, batches), a
shared group offset $b_{g(i)} \sim N(0, \tau^2)$ can be added to every
component's linear predictor with `random = ~ group`:

```{r re, eval = FALSE}
# grouped data with 12 groups; component slopes +2 / -2
reg <- nimixReg(y ~ x, df, random = ~ g, K = 2, method = "fixedk",
                mcmcControl = list(niter = 3000, nburnin = 1200), seed = 1)
sort(relabel(reg)@relabeled$summary$x)
#> [1] -1.93  1.97                    # slopes recovered with the offset present
mean(reg@mcmcSamples[, "tauRE"])       # the between-group SD, tau
```

Two things are worth knowing. The offsets use a **sum-to-zero**
parameterisation, because with free $b$ the component intercepts and the mean
of $b$ form a translation ridge that mixes badly (a prototype measured
`cor` = -0.979 and a minimum ESS of 25 in 2500 draws; the constraint restored
it to ~205). So the reported `b` are centred, and the component intercepts
absorb the group mean. And the exact conjugate coefficient sampler is
extended to condition on the current offsets, which keeps the fit
scale-equivariant in the predictors even with the random effect active.

Current scope: `method = "fixedk"`, `distribution = "normal"`, one grouping
factor, random intercept. Random slopes and further families follow the
gated roadmap.
