---
title: "Heavy tails: Student-t vs Normal-Gamma components"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Heavy tails: Student-t vs Normal-Gamma components}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

nimix offers two mathematically equivalent heavy-tailed component families:

* `distribution = "studentt"` -- the Student-t density used **directly**
  (non-conjugate; slice sampling on the component parameters);
* `distribution = "normalgamma"` -- the same marginal Student-t obtained by
  **data augmentation** (`y_i | omega_i ~ N(mu_k, sigma_k^2 / omega_i)`,
  `omega_i ~ Gamma(df/2, df/2)`), which restores conjugacy at the price of one
  latent scale per observation.

The marginal model is identical; the *computational* behaviour is not, and
that trade-off is the point of this vignette. We compare both on official
statistics: log GDP per capita from `wdi2022` (`?wdi2022`), a skewed,
heavy-tailed cross-country distribution.

> MCMC chunks use `eval = FALSE` (CRAN time limits); printed results are from
> an actual run on this data.

## Data

```{r data, eval = TRUE}
library(nimix)
data(wdi2022)
ly <- log(wdi2022$gdp_pc)
```

## Both fits

```{r fits}
ctrl <- list(niter = 4000, nburnin = 1500)
ft <- relabel(nimixClust(ly, K_max = 8, distribution = "studentt",
                         mcmcControl = ctrl, seed = 1))
fg <- relabel(nimixClust(ly, K_max = 8, distribution = "normalgamma",
                         mcmcControl = ctrl, seed = 1))
summary(ft); summary(fg)
```

Actual results on this dataset:

```
#> studentt     : modalK = 2, regimes at log-GDP 8.00 / 10.06,
#>                ESS(#clusters) = 147, ~30 s
#> normalgamma  : modalK = 3, dominant regimes 8.07 / 10.10 (+ small 10%
#>                middle component), ESS(#clusters) = 116, ~38 s
```

Both agree on the two dominant income regimes. The augmented path found one
extra small component and mixed somewhat more slowly on the partition.

## Which one should you use?

The augmentation restores conjugacy, but the per-observation `omega_i` couple
tightly to the partition, which can slow the mixing of the *number of
clusters* -- a penalty we found consistently in benchmarks and mitigate by
replacing the default random-walk sampler on `omega` with slice sampling
(done automatically via the spec's sampler hook). Practical guidance:

* `"studentt"` is the sensible default for heavy-tailed clustering;
* `"normalgamma"` is worth trying when its Gibbs updates are advantageous
  (e.g. larger n per component), and its `omega_i` posteriors are themselves
  useful: small values flag the observations the model treats as outliers
  (a robustness-weight interpretation).

Always compare `summary()` ESS diagnostics on your own data rather than
assuming one is uniformly better.
