---
title: "Skew multivariate clustering with an estimated orthogonal factor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Skew multivariate clustering with an estimated orthogonal factor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Real multivariate clusters are often skewed, and not along the coordinate
axes. `nimix` implements the Ferreira & Steel (2007) skew multivariate
construction, in which each component is

$$\eta = A^\top \epsilon + \mu, \qquad A = OU,$$

with $\epsilon$ having independent Fernandez-Steel-skewed margins, $U$
the upper triangle of $\Sigma = A^\top A$, and $O$ an orthogonal
factor that rotates the skew directions away from the axes. When $O$ is
estimated, the model can point its asymmetry in any direction.

Two component families are available -- skew-Normal and skew-independent-
Student -- each in a fixed-$O$ and an estimated-$O$ variant. This
article uses the estimated-$O$ skew-Normal in three dimensions.

> MCMC chunks use `eval = FALSE` (CRAN time limits); the printed results are
> from an actual run.

## Simulate two skewed, rotated clusters

```{r}
library(nimix)

rgen <- function(n, mu, Sigma, gamma, theta) {
  m <- length(mu)
  O <- orthogonalFactor(theta, m)          # Householder factor from angles
  U <- chol(Sigma)
  G <- matrix(gamma, n, m, byrow = TRUE)
  W <- abs(matrix(rnorm(n * m), n, m))
  eps <- ifelse(matrix(runif(n * m), n, m) < G^2 / (1 + G^2), W * G, -W / G)
  sweep(eps %*% O %*% U, 2, mu, "+")
}

set.seed(40)
Sigma <- diag(3); Sigma[1, 2] <- Sigma[2, 1] <- 0.25
Y <- rbind(rgen(250, c(-4, -4, -4), Sigma, c(0.5, 2.0, 1.5), c( 0.30, -0.20,  0.40)),
           rgen(250, c( 4,  4,  4), Sigma, c(2.0, 0.5, 0.7), c(-0.25,  0.35, -0.30)))
```

## Fit

The user-facing name is `"skewnormal-mv-o"` regardless of dimension; the fit
routes on the data's column count. With two columns it uses a dedicated
bivariate implementation; with more, a general Householder parameterisation
with $m(m-1)/2$ angles.

```{r}
fit <- relabel(nimixClust(Y, K = 2, method = "fixedk",
                          distribution = "skewnormal-mv-o",
                          mcmcControl = list(niter = 2500, nburnin = 1000),
                          seed = 3))
class(fit@distSpec)
#> [1] "SkewNormalMvOGenSpec"
```

Partition and location recovery are exact:

```{r}
s <- fit@relabeled$summary
sort(c(s$mu_1_mean))
#> [1] -3.85  4.21           # true -4, 4
```

## Reading gamma and O: canonicalisation

Here is the one thing that trips people up. The rows of $A$ can be
permuted and sign-flipped without changing the density -- this is label
switching, but in the *dimension* index rather than the component index.
`nimix` resolves it by mapping every posterior draw to a unique canonical
representative (Ferreira & Steel's identifiability restriction, applied
*post hoc* rather than as a sampling constraint). So the reported `gamma` and
`O` are **canonicalised**, and to compare them with the simulating values you
must canonicalise those too:

```{r}
c1 <- canonicaliseO(orthogonalFactor(c(0.30, -0.20, 0.40), 3), c(0.5, 2.0, 1.5))
round(c1$gamma, 2)
#> [1] 0.50 0.50 0.67          # canonicalised truth, NOT the raw c(0.5, 2.0, 1.5)

round(c(s$gamma_1_mean[1], s$gamma_2_mean[1], s$gamma_3_mean[1]), 2)
#> [1] 0.49 0.57 0.65          # estimate -- matches the canonicalised truth
```

Every draw found a canonical representative:

```{r}
fit@relabeled$canonicalFraction
#> [1] 1
```

## Why restriction (8) is a canonicalisation, not a constraint

Ferreira & Steel suggest confining the Householder angles to a box so that
$O$ is identified. Testing that directly shows it does not work as a
sampling constraint -- only a small fraction of angle draws from the box
actually satisfy the identifiability restriction, and the fraction shrinks
fast with dimension:

```{r}
frac <- mean(replicate(500, {
  th <- runif(3, -pi/2, pi/2)
  nimix:::.restriction8(orthogonalFactor(th, 3))
}))
frac
#> [1] 0.068                   # ~7% at m = 3; ~0.7% at m = 4
```

Constraining a sampler to a 7% (or 0.7%) slice of its own prior would mix
badly. Applying the restriction as a post-hoc canonicalisation instead --
exactly one signed row permutation satisfies it -- keeps the sampler
unconstrained and the density invariant.

## Caveats for the general-m families

The estimated-$O$ families with $m > 2$ are marked **experimental**.
Partition and location recovery are robust, but:

- `gamma` and `O` are reported *after* canonicalisation (compare against
  canonicalised truth, as above);
- `O_mean` is an elementwise posterior mean and is **not** itself orthogonal
  -- use the `O` draws when orthogonality matters;
- individual angles are large-sample quantities, and the mirror modes of the
  angle likelihood multiply with `m`; budget more iterations than for the
  fixed-$O$ families, especially under the MRF engine.

The heavy-tailed sibling, `"skewistudent-mv-o"`, follows the identical
workflow with an extra per-margin degrees-of-freedom parameter `nu`.

## References

Ferreira, J. T. A. S. & Steel, M. F. J. (2007). A new class of skewed
multivariate distributions with applications to regression analysis.
*Statistica Sinica* 17, 505--529.
