---
title: "Univariate mixture clustering with the DPM engine"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{Univariate mixture clustering with the DPM engine}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette demonstrates the v0.1.0 functionality of **nimix**: Bayesian
univariate Gaussian mixture clustering using a Dirichlet Process Mixture (DPM)
engine built on NIMBLE's Chinese Restaurant Process (`dCRP`).

> Chunks use `eval = FALSE` so the vignette builds quickly under CRAN time
> limits . Run them interactively.

## Fit

```{r}
library(nimix)

set.seed(1)
y <- c(rnorm(150, -4, 1), rnorm(150, 4, 1)) # two clusters, K_true = 2

fit <- nimixClust(
 y, K_max = 10, distribution = "normal", method = "dpm",
 mcmcControl = list(niter = 6000, nburnin = 2000)
)
fit
```

The DPM does not fix the number of clusters: it places a Chinese Restaurant
Process prior on the allocation vector and lets the data determine how many
components are occupied. The truncation level `K_max` is the length of the
cluster-parameter vectors, **not** a fixed number of clusters.

## Label switching: relabel before summarising

The mixture likelihood is invariant to permuting component labels, so raw
per-component posterior means are meaningless. `summary` therefore relabels
first (ECR-ITERATIVE-1 from the `label.switching` package), conditioning on
the modal number of occupied clusters.

```{r}
summary(fit)
```

You can see the switching directly:

```{r}
plot(fit, type = "trace_raw") # zig-zags between levels = switching
plot(fit, type = "trace_relabeled") # stable bands after relabelling
```

## Posterior of the number of clusters

```{r}
plot(fit, type = "K")
```

## Posterior predictive density

```{r}
plot(fit, type = "density")
pp <- predict(fit, newdata = seq(-8, 8, length.out = 200))
head(pp)
```

## Choosing `K_max` and the concentration prior

`K_max` is a truncation, not the answer. A useful default is
`min(10, floor(n / 5))`. Posterior conclusions about K depend on the prior for
the concentration parameter $\alpha$; nimix places a Gamma hyperprior on
$\alpha$ (configurable via `prior = list(concPrior = c(shape, rate))`) so the
data inform the clustering granularity rather than a fixed value
.

## References

- de Valpine et al. (2017), *JCGS* 26(2), 403–413.
- Neal (2000), *JCGS* 9(2), 249–265.
- Papastamoulis & Iliopoulos (2010), *JCGS* 19(2), 313–331.
- Papastamoulis (2016), *JSS Code Snippets* 69(1).
