---
title: "Multivariate mixture clustering with the DPM engine"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{Multivariate mixture clustering with the DPM engine}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette demonstrates the v0.2.0 functionality of **nimix**: Bayesian
**multivariate** Gaussian mixture clustering on the same Dirichlet Process
Mixture (DPM) engine used for the univariate case. The only thing that changes
from the user's point of view is that `data` is now a numeric **matrix** (one
row per observation, one column per dimension).

> Chunks use `eval = FALSE` so the vignette builds quickly under CRAN time
> limits . Run them interactively.

## The model

Each component is a multivariate normal. The conjugate cluster base measure is
a **Normal-Inverse-Wishart**:
$$
\Sigma_j \sim \text{InvWishart}(S_0, df_0), \qquad
\mu_j \mid \Sigma_j \sim N_d(\mu_0, \Sigma_j / \kappa_0),
$$
with the allocation vector following a Chinese Restaurant Process. nimix writes
this with NIMBLE's `dmnorm(cov = ...)` + `dinvwish`, which is the pairing NIMBLE
recognises as conjugate for `dCRP` — so the efficient collapsed CRP samplers are
used automatically . The defaults are
*data-scaled* : `mu0 = colMeans(data)`, the prior
mean of the cluster covariance equals `cov(data)`, and `df0 = d + 2` (validated
to exceed `d + 1` so empty-component draws stay non-singular).

## Fit

```{r}
library(nimix)

set.seed(1)
# two well-separated 2-D clusters, K_true = 2
Y <- rbind(
 matrix(rnorm(150 * 2, mean = -3), ncol = 2),
 matrix(rnorm(150 * 2, mean = 3), ncol = 2)
)

fit <- nimixClust(
 Y, K_max = 10, distribution = "normal", method = "dpm",
 mcmcControl = list(niter = 6000, nburnin = 2000)
)
fit
```

`nimixClust` detects that `Y` is a matrix and routes it to the multivariate
component (`NormalMvSpec`). You can force the family with
`distribution = "normal-mv"` (or `"normal-uv"` for a vector).

## Label switching: relabel before summarising

As in the univariate case, the mixture likelihood is invariant to permuting
component labels, so `summary` relabels first (ECR-ITERATIVE-1), conditioning
on the modal number of occupied clusters. For multivariate fits the component
table reports the posterior mean of each mean coordinate (`mu_1`, `mu_2`, ...)
and the posterior mean of each marginal variance (`var_1`, `var_2`, ...); the
full relabelled mean covariance matrices are available in
`fit@relabeled$Sigma_mean`.

```{r}
summary(fit)
```

## Visualising the clusters

```{r}
plot(fit, type = "K") # posterior of the number of occupied clusters
plot(fit, type = "cluster") # data coloured by MAP cluster (first two dims)
```

The label-switching diagnostics also work in the multivariate case (they plot
the first mean coordinate across clusters):

```{r}
plot(fit, type = "trace_raw")
plot(fit, type = "trace_relabeled")
```

## Posterior predictive density at points

For multivariate data the predictive density is evaluated at supplied rows
rather than over a 1-D grid:

```{r}
newpts <- rbind(c(-3, -3), c(0, 0), c(3, 3))
predict(fit, newdata = newpts)
```

## Notes on dimension and `K_max`

* `K_max` remains a truncation level, not a fixed number of clusters.
* The prior covariance scaling matters more in higher dimensions; the
 data-scaled Normal-Inverse-Wishart default keeps the prior weakly informative
 without becoming vague .
* With small `n` relative to `d`, identifiability weakens (
); prefer `n` comfortably larger than `d` and inspect the posterior
 of K rather than trusting a single modal value.

## References

- de Valpine et al. (2017), *JCGS* 26(2), 403–413.
- Neal (2000), *JCGS* 9(2), 249–265.
- Zhang, Chan, Wu & Chen (2004), *Statistics and Computing* 14, 343–355.
- Dellaportas & Papageorgiou (2006), *Statistics and Computing* 16, 57–68.
- Papastamoulis & Iliopoulos (2010), *JCGS* 19(2), 313–331.
